Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-22 Paper/method.tex

In this section, we present the computational strategy used for computing the
active subspace in the case of H$_2$/O$_2$ reaction kinetics with uncertain
pre-exponent of the rate law associated with individual reactions provided in
Table~\ref{tab:kinetics}.
Specifically, we employ two strategies: 
A gradient-based strategy that involves computation of model
gradients in order to construct the matrix $\hat{\mat{C}}$
in~\eqref{eq:chat}, and a gradient-free approach 
that involves a regression-based local linear approximation of
the model output to circumvent the computational cost associated with gradient
computation. 
The gradient-based approach is based on Algorithm
1.1, and the gradient-free approach is based on Algorithm 1.2
in~\cite{Constantine:2015}. The two strategies are implemented in an
iterative manner to ensure computational efficiency by avoiding 
extraneous model evaluations once a converged active subspace
is achieved.

@@

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_0$ samples denoted by $\bm{\xi}_i$, $i$ = $1,\ldots,n_0$.
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and the set of corresponding eigenvalues, 
$\hat{\mat{\Lambda}}_1$.
 At each subsequent iteration, model evaluations are
generated at a new set of $n_k$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to have converged once
the maximum relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$
($j$ is used as an index for the eigenvectors),
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide an insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

	%	\State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$.%
		\State Set $N_\text{total}$ = $N_\text{total}$ + $n_k$ 
		\State Compute $\bm{g}^i = \nabla_{\bm{\xi}_i}f(\bm{\xi}_i)$, 
             	$i=n_{k-1}+1, \ldots, n_{k-1}+n_k$.  
		\State Compute $\hat{\mat{C}}$ = 
        	$\frac{1}{N_\text{total}}\sum\limits_{k=1}^{N_\text{total}}[\bm{g}^i][\bm{g}^i]^\top$
		\State Eigenvalue decomposition, $\hat{\mat{C}}$ = $\hat{\mat{W}}^{(k)}\hat{\mat{\Lambda}}^{(k)}
		 \hat{\mat{W}}^{(k)\top}$
		\State Compute $\delta \hat{\mat{W}}_{1,j}^{(k)}$ = 
                       \scalebox{1.25}{$\frac{\|(\hat{\mat{W}}_{1,j}^{k})^2 - 
                       (\hat{\mat{W}}_{1,j}^{k-1})^2\|}{\|(\hat{\mat{W}}_{1,j}^{k-1})^2\|}$}, 
                       $j = 1,\ldots,r$.
		\If {$\max\left(\delta \hat{\mat{W}}_{1,j}^{(k)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\nu_{i,r}(f) = \sum\limits_{j=1}^{r} \hat{\mat{\Lambda}}_{i,j}\hat{\mat{W}}_{i,j}^2$,
	$i=1,\ldots,N_p$.
	\State Normalize $\nu_{i,r}(f)$ as $\tilde{\nu_{i,r}}(f)$ = \scalebox{1.25}{$\frac{\nu_{i,r}(f)}{\sum_i\nu_{i,r}(f)}$}.

@@

To motivate our iterative strategy, we implement
Algorithm~\ref{alg:grad} to the 19-dimensional H$_2$/O$_2$ reaction kinetics
problem with uncertain $A_i$'s. For the purpose of verification,
$\hat{\mat{C}}$ is initially constructed using a large set of samples
($N$~=~1000) in the input domain and the corresponding active subspace is
computed. In Figure~\ref{fig:eig_comp}~(left), we illustrate the comparison of
the resulting normalized eigenvalue spectrum ($\lambda_1, \ldots, 
\lambda_{19}$) with the same using a much 
smaller set of samples, $n$~=~$\{20,40,80,120\}$.

%Additionally, we compare relative L-2
%norm of the difference between the squared value of corresponding components
%of the eigenvectors, obtained using the large set, $N$, and the smaller sets, $n$.
%In  Figure~\ref{fig:eig_comp} (right), the relative L-2 error norm for the
%first five eigenvectors is plotted.

%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/eig_comp}
   \includegraphics[width=0.48\textwidth]{./Figures/err_eigv_1}
\caption{Left: A comparison of the normalized eigenvalue spectrum using $n$ = $\{20,40,80,120\}$ samples with that
obtained using a much larger sample size, $N$~=~1000. Right: Relative $L_2$ 
norm of the difference between
the square of individual components ($w_i$'s) of the dominant eigenvector evaluated using $N$=1000 and $n$ = $\{20,40,80,120\}$ samples.} 
\label{fig:eig_comp}
\end{center}
\end{figure}
%
We observe that the dominant eigenvalues, $\lambda_1, \ldots, \lambda_4$, 
are approximated 
reasonably well with just 20 samples. As expected, the accuracy of higher-order eigenvalues is observed
to improve with the sample size. Since 
$\left(\frac{\lambda_1}{\lambda_2}\right)\gg 1$, we expect 
a 1-dimensional active subspace. Hence, the first eigenvector 
sufficiently captures the uncertainty in
the  model output. To confirm this, we plot a relative L-2 norm of the difference between the 
squared value of corresponding components of the dominant eigenvector, computed using $N$ = 1000
and $n$ = $\{20,40,80,120\}$ samples in Figure~\ref{fig:eig_comp} (right). 
The relative L-2 error norm is found to be $\mathcal{O}(10^{-2})$ even for the case, $n$ = 20.
Hence, a sample size of 20 is adequate for computing the active subspace in this
case. The iterative strategy therefore offers a significant potential for computational advantage. 
In the following section, we provide a brief description of the gradient-free approach and compare
numerical results for the active subspace and the global sensitivity measures based on the two
approaches.

@@

%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/comp_as}
  \includegraphics[width=0.45\textwidth]{./Figures/ub_conv_kinetics_rich}
\caption{Left: A bar-graph of normalized activity scores ($\tilde{\nu_{i,r}}$'s) for the 19 uncertain pre-exponents ($A_i$'s); $r$
denotes the number of eigenvectors in the active subspace.
Right: A plot of the screening metric ($\widehat{\mathcal{C}_i\mu_i}$), defined as a normalized product of the
Poincare constant and the DGSM ($\mu_i$) (adapted from~\cite{Vohra:2018}).}
\label{fig:comp_as}
\end{center}
\end{figure}
%
The activity scores estimated using the two approaches agree favorably with each other as well as those
based on the screening metric involving the DGSMs from~\cite{Vohra:2018}. It is observed that the uncertainty
associated with the ignition delay is largely due to the uncertainty in $A_9$ and $A_1$. Sensitivity towards
$A_{15}$ and $A_{17}$ is also found to be significant.

>>>>

In this section, we present a generalized framework for computing the
active subspace. The framework is employed to the H$_2$/O$_2$ reaction kinetics
problem whereby the 
pre-exponent ($A_i$) in the rate law associated with individual reactions provided in
Table~\ref{tab:kinetics} is considered to be uniformly
distributed in the interval, $[0.9A_i^\ast, 1.1A_i^\ast]$; $A_i^\ast$ is the nominal
estimate provided in~\cite{Yetter:1991}.
Specifically, we employ two strategies: 
A gradient-based strategy that involves computation of model
gradients in order to construct the matrix $\hat{\mat{C}}$
in~\eqref{eq:chat}, and a gradient-free approach 
that involves a regression-based local linear approximation of
the model output to circumvent the computational cost associated with gradient
computation. 
The gradient-based approach is based on Algorithm
1.1, and the gradient-free approach is based on Algorithm 1.2
in~\cite{Constantine:2015}. The two strategies are implemented in an
iterative manner to ensure computational efficiency by avoiding 
extraneous model evaluations once a converged active subspace
is achieved.

@@

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_0$ samples denoted by $\bm{\xi}_i$, $i$ = $1,\ldots,n_0$.
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and the set of corresponding eigenvalues, 
$\hat{\mat{\Lambda}}_1$. Note that $\hat{\mat{W}}_1$ is obtained by
partitioning the eigenspace around $\lambda_j$ such that the ratio of
subsequent eigenvalues,
$\left(\frac{\lambda_j}{\lambda_{j+1}}\right)>\mathcal{O}(10^2)$.
 At each subsequent iteration, model evaluations are
generated at a new set of $n_k$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to have converged once
the maximum relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$
($j$ is used as an index for the eigenvectors),
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide an insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

	%	\State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$.%
		\State Set $N_\text{total}$ = $N_\text{total}$ + $n_k$ 
		\State Compute $\bm{g}^i = \nabla_{\bm{\xi}_i}f(\bm{\xi}_i)$, 
             	$i=n_{k-1}+1, \ldots, n_{k-1}+n_k$.  
		\State Compute $\hat{\mat{C}}$ = 
        	$\frac{1}{N_\text{total}}\sum\limits_{k=1}^{N_\text{total}}[\bm{g}^i][\bm{g}^i]^\top$
		\State Eigenvalue decomposition, $\hat{\mat{C}}$ = $\hat{\mat{W}}^{(k)}\hat{\mat{\Lambda}}^{(k)}
		 \hat{\mat{W}}^{(k)\top}$
		\State Compute $\delta \hat{\mat{W}}_{1,j}^{(k)}$ = 
                       \scalebox{1.25}{$\frac{\|(\hat{\mat{W}}_{1,j}^{k})^2 - 
                       (\hat{\mat{W}}_{1,j}^{k-1})^2\|}{\|(\hat{\mat{W}}_{1,j}^{k-1})^2\|}$}, 
                       $j = 1,\ldots,r$.
		\If {$\max\left(\delta \hat{\mat{W}}_{1,j}^{(k)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\nu_{i,r}(f) = \sum\limits_{j=1}^{r} \Lambda_j w_{i,j}^2$,
	$i=1,\ldots,N_p$.
	\State Normalize $\nu_{i,r}(f)$ as $\tilde{\nu}_{i,r}(f)$ = \scalebox{1.25}{$\frac{\nu_{i,r}(f)}{\sum_i\nu_{i,r}(f)}$}.

@@

To motivate our iterative strategy, we implement
Algorithm~\ref{alg:grad} to the 19-dimensional H$_2$/O$_2$ reaction kinetics
problem with uncertain $A_i$'s. For the purpose of verification,
$\hat{\mat{C}}$ is initially constructed using a large set of samples
($N$~=~1000) in the input domain and the corresponding active subspace is
computed. In Figure~\ref{fig:eig_comp}, we illustrate the comparison of
the resulting normalized eigenvalue spectrum ($\lambda_1, \ldots, 
\lambda_{19}$) with the same using a much 
smaller set of samples, $n$~=~$\{20,40,80,120\}$.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/eig_comp}
%   \includegraphics[width=0.48\textwidth]{./Figures/err_eigv_1}
\caption{A comparison of the normalized eigenvalue spectrum using $n$ = $\{20,40,80,120\}$ samples with that
obtained using a much larger sample size, $N$~=~1000. 
%Right: Relative $L_2$ 
%norm of the difference between
%the square of individual components ($w_i$'s) of the dominant eigenvector evaluated using 
$N$=1000 and $n$ = $\{20,40,80,120\}$ samples.
} 
\label{fig:eig_comp}
\end{center}
\end{figure}
%
We observe that the dominant eigenvalues, $\lambda_1, \ldots, \lambda_4$, 
are approximated 
reasonably well with just 20 samples. As expected, the accuracy of higher-order eigenvalues is observed
to improve with the sample size. Since 
$\left(\frac{\lambda_1}{\lambda_2}\right)\gg 1$, we expect 
a 1-dimensional active subspace. Hence, the first eigenvector 
sufficiently captures the uncertainty in
the  model output. To further confirm this, we evaluate a relative L-2 norm of the difference
 ($\varepsilon_\text{L-2}^{N-n}$) between the 
squared value of corresponding components of the dominant eigenvector, computed using $N$ = 1000
and $n$ = $\{20,40,80,120\}$.  The quantity, $\varepsilon_\text{L-2}^{N-n}$, was found to be 
$\mathcal{O}(10^{-2})$ for $n$ = 20, 40, and 80; and $\mathcal{O}(10^{-3})$ for $n$ = 120.
Thus, even a small sample size, $n$ = 20, seems to approximate the dominant eigenspace with
reasonable accuracy in this case. 
The iterative strategy therefore offers a significant potential for computational advantage. 
In the following section, we provide a brief description of the gradient-free approach and compare
numerical results for the active subspace and the global sensitivity measures based on the two
approaches.

@@

%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/comp_as}
  \includegraphics[width=0.45\textwidth]{./Figures/ub_conv_kinetics_rich}
\caption{Left: A bar-graph of normalized activity scores ($\tilde{\nu_{i,r}}$'s) 
for the 19 uncertain pre-exponents ($A_i$'s); $r$
denotes the number of eigenvectors in the active subspace.
Right: A plot of the screening metric ($\tilde{\nu}_i$) for individual $A_i$'s 
(adapted from~\cite{Vohra:2018}).}
\label{fig:comp_as}
\end{center}
\end{figure}
%
The activity scores estimated using the two approaches agree favorably with each other as well as those
based on the screening metric involving the DGSMs from~\cite{Vohra:2018}. It is observed that the uncertainty
associated with the ignition delay is largely due to the uncertainty in $A_9$ and $A_1$. Sensitivity towards
$A_{15}$ and $A_{17}$ is also found to be significant.
